# Overview {#overview}

## Aims

This module aims to provides students with a range of techniques for analysing and modelling spatial data:

* build upon the more general research training delivered via companion modules on *Data Collection and Data Analysis*, both of which have an aspatial focus;
* highlight a number of key social issues that have a spatial dimension;
* explain the specific challenges faced when attempting to analyse spatial data;
* introduce a range of analytical techniques and approaches suitable for the analysis of spatial data; and,
* enhance practical skills in using *R* software packages to implement a wide range of spatial analytical tools.

## Learning Outcomes

By the end of the module, students should be able to:

* identify some key sources of spatial data and resources of spatial analysis and modelling tools;
* explain the advantages of taking spatial structure into account when analysing spatial data;
* apply a range of computer-based techniques for the analysis of spatial data, including mapping, correlation, kernel density estimation, regression, multi-level models, geographically-weighted regression, spatial interaction models and spatial econometrics;
* apply appropriate analytical strategies to tackle the key methodological challenges facing spatial analysis – spatial autocorrelation, heterogeneity, and ecological fallacy; and, 
* select appropriate analytical tools for analysing specific spatial data sets to address emerging social issues facing the society.



## Computational Environment

Dependencies + Landing page for guides

### Dependency list

List of libraries used in this book:

```{r, message = FALSE, warning = FALSE}
deps <- list(
    "arm",
    "car",
    "corrplot",
    "FRK",
    "gghighlight",
    "ggplot2",
    "ggmap",
    "GISTools",
    "gstat",
    "jtools",
    "kableExtra",
    "knitr",
    "lme4",
    "lmtest",
    "lubridate",
    "MASS",
    "merTools",
    "plyr",
    "RColorBrewer",
    "rgdal",
    "sf",
    "sjPlot",
    "sp",
    "spgwr",
    "spatialreg",
    "spacetime",
    "stargazer",
    "tidyverse",
    "tmap",
    "viridis"
)
```

And we can load them all to make sure they are installed:

```{r}
for(lib in deps){library(lib, character.only = TRUE)}
```